# Common Functions {#sec-common-functions}

This chapter covers the small set of functions that will carry the majority of your day-to-day work. Each one maps to something you already do in Excel, SAS, or Access — the goal here is naming the R equivalent and giving you a working recipe, not re-teaching the underlying logic you already have.

All examples continue using `static_file`, loaded back in [Your First Import](#sec-first-import).

## Cleaning messy column names first

Real files rarely arrive with clean, consistent column names. Before doing anything else, it's worth knowing about `janitor::clean_names()` — genuinely one of the highest-value, lowest-effort functions in this entire book:

```{r}
#| include: false
#| message: false
library(tidyverse)

static_file <- read_csv("data/static_file_sample.csv")
```

```{r}
library(janitor)

static_file_new_names <- static_file |> clean_names()

names(static_file_new_names)[1:10]
```

This converts inconsistent capitalization, spaces, and punctuation in column names into a single consistent format (typically `snake_case`) automatically. Our Static File sample already has clean names, so you won't see dramatic changes here — but the moment you load a real Excel export with column headers like `"School Name "` (trailing space included) or `"% Proficient"`, this single line will save you real frustration. Use it by habit on every new file you import.

## select() — choosing columns

The equivalent of hiding columns in Excel, except explicit about what you're keeping rather than what you're hiding:

```{r}
static_file |> select(SchoolName, DistrictName, Subject, Performance)

static_file |> select(-FirstName, -MiddleName, -LastName)   # drop specific columns, keep the rest

static_file |> select(starts_with("EL"))   # all EL-related columns at once
```

## filter() — choosing rows

The equivalent of AutoFilter, except it returns a clean result rather than just hiding rows visually:

```{r}
static_file |> filter(Subject == "Math")

static_file |> filter(ScaleScore >= 450 & Subject == "ELA")

static_file |> filter(County %in% c("Maricopa", "Pima"))
```

## mutate() — creating or changing a column

The equivalent of adding a formula column, applied automatically to every row (recall vectorized behavior from [Fundamental Concepts](#sec-fundamental-concepts)):

```{r}
static_file |>
  mutate(scale_score_rounded = round(ScaleScore))
```

## if_else() and case_when() — conditional logic

`if_else()` handles a single yes/no condition — the direct equivalent of Excel's `IF()`:

```{r}
static_file |>
  mutate(is_high_school = if_else(StudentGrade %in% c("9","10","11","12"), "High School", "Not High School"))
```

`case_when()` handles multiple conditions in sequence — the equivalent of a nested `IF()` chain, but far easier to read and far harder to make a logic error in, since it's evaluated top to bottom and stops at the first matching condition:

```{r}
static_file |>
  mutate(performance_label = case_when(
    Performance == 1 ~ "Minimally Proficient",
    Performance == 2 ~ "Partially Proficient",
    Performance == 3 ~ "Proficient",
    Performance == 4 ~ "Highly Proficient",
    TRUE              ~ NA_character_
  ))
```

The final `TRUE ~ ...` line is your catch-all — anything that didn't match an earlier condition lands here. We've set it to `NA_character_` (an explicitly-typed missing text value) since `Performance` is `NA` for students who weren't assessed, and we want that to flow through cleanly rather than becoming a stray label.

## arrange() — sorting

```{r}
static_file |> arrange(DistrictName, desc(ScaleScore))
```

Ascending by default; wrap a column in `desc()` to reverse just that one — useful when, as above, you want districts alphabetical but scores highest-first within each.

## rename() — renaming columns

```{r}
static_file |> rename(school = SchoolName, district = DistrictName)
```

## count() — quick frequency tables

```{r}
static_file |> count(Subject)

static_file |> count(County, Charter)   # cross-tab style: every County x Charter combination
```

## str_detect() — pattern matching inside text

Part of the `stringr` package (included in the tidyverse), useful for searching within a text column — similar in spirit to Excel's `SEARCH()` or a SQL `LIKE` clause:

```{r}
static_file |> filter(str_detect(SchoolName, "Online"))
```

## round() — controlling decimal precision

```{r}
static_file |> mutate(scale_score_rounded = round(ScaleScore, 1))   # one decimal place
```

## Putting several together

This is where the pipe (`|>`) starts to earn its keep — chaining several of the functions above into a single readable sequence:

```{r}
static_file |>
  filter(Subject == "ELA", !is.na(Performance)) |>
  select(SchoolName, DistrictName, Performance) |>
  mutate(performance_label = case_when(
    Performance == 1 ~ "Minimally Proficient",
    Performance == 2 ~ "Partially Proficient",
    Performance == 3 ~ "Proficient",
    Performance == 4 ~ "Highly Proficient"
  )) |>
  arrange(DistrictName)
```

Read top to bottom as one sentence: *take the Static File, keep only assessed ELA records, keep just these columns, add a readable label, then sort by district.* Each line is a single, understandable step — and the whole sequence reads roughly the way you'd describe the task out loud to a colleague.

The next chapter builds directly on this same pattern to tackle the task you'll do more than almost any other in accountability work: summarizing data by group.
