# Visualization {#sec-visualization}

## The grammar of graphics, in plain terms

Excel's Chart Wizard asks you one question up front: *what type of chart do you want?* `ggplot2` — the charting package you'll use throughout this chapter — asks a different question first: ***what does each variable in your data mean in the picture you're about to build?*** This way of thinking is called the **grammar of graphics**, and once it clicks, it tends to make charting in R feel less like memorizing chart types and more like describing what you already understand about your data.

Two pieces of vocabulary worth having before the first example:

- **Aesthetics (`aes()`)** — which columns map to which visual properties. Does `ScaleScore` control position on the x-axis? Does `Subject` control color? You're deciding what each variable's *job* is in the picture, before deciding how anything gets drawn.
- **Layers** — think of a chart as a stack of transparent sheets, each one adding something: one layer draws points, another adds a trend line, another adds labels. You build a chart by stacking layers with `+`, and you can add, remove, or reorder them independently.

## The basic pattern

```{r}
#| eval: false
library(ggplot2)

ggplot(data = static_file, aes(x = ScaleScore, y = SGP_CCR)) +
  geom_point()
```

Read this as a sentence: *"Using static_file, map ScaleScore to the x-axis and SGP_CCR to the y-axis, then draw points."* `aes()` set up the meaning; `geom_point()` is the layer that actually draws something using that meaning.

## Common chart types — same data, different geom

The aesthetics often barely change between chart types — usually, only the `geom_*()` layer does:

```{r}
#| eval: false
static_file |>
  filter(!is.na(Performance)) |>
  count(Subject, Performance) |>
  ggplot(aes(x = factor(Performance), y = n, fill = Subject)) +
  geom_col(position = "dodge")             # bar chart

ggplot(static_file, aes(x = ScaleScore)) +
  geom_histogram()                          # histogram

ggplot(static_file, aes(x = factor(StudentGrade), y = ScaleScore)) +
  geom_boxplot()                            # boxplot, score distribution by grade
```

## Adding color, labels, and a cleaner look

```{r}
#| eval: false
static_file |>
  filter(!is.na(Performance), Subject == "ELA") |>
  ggplot(aes(x = ScaleScore, fill = factor(Performance))) +
  geom_histogram(alpha = 0.8) +
  labs(
    title = "ELA Scale Score Distribution by Performance Level",
    x = "Scale Score",
    y = "Number of Students",
    fill = "Performance Level"
  ) +
  theme_minimal()
```

`labs()` is where your titles and axis labels live; `theme_minimal()` is one of several built-in themes that strip away Excel's default gray gridline clutter. `theme_bw()`, `theme_classic()`, and others are worth trying — swapping a theme is a one-word change, not a re-build.

## Faceting — one chart per group, automatically

```{r}
#| eval: false
static_file |>
  filter(!is.na(ScaleScore)) |>
  ggplot(aes(x = ScaleScore)) +
  geom_histogram() +
  facet_wrap(~ Subject)
```

This single line replaces manually building and formatting a separate chart per subject — `facet_wrap()` does it in one pass, with consistent formatting guaranteed across every panel.

## Making it interactive: plotly::ggplotly()

Everything above produces a static image. Wrapping a finished `ggplot2` chart in `ggplotly()` turns it interactive — hover tooltips, zoom, and pan — with no rebuilding required:

```{r}
#| eval: false
library(plotly)

p <- static_file |>
  filter(!is.na(Performance)) |>
  ggplot(aes(x = factor(Performance), fill = Subject)) +
  geom_bar(position = "dodge") +
  labs(title = "Student Counts by Performance Level and Subject")

ggplotly(p)
```

This is particularly effective inside a Quarto HTML document or presentation (see [Project Walkthrough 2](#sec-project-2)) — a static chart in a PDF becomes something a viewer can actually explore.

## Tables: gt for polished static tables

The `gt` package ("grammar of tables," deliberately named to echo `ggplot2`'s philosophy) turns a summarized data frame into a publication-quality formatted table:

```{r}
#| eval: false
library(gt)

static_file |>
  group_by(SchoolName) |>
  summarize(n_students = n(), avg_scale_score = mean(ScaleScore, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(avg_scale_score)) |>
  head(10) |>
  gt() |>
  tab_header(title = "Top 10 Schools by Average Scale Score") |>
  fmt_number(columns = avg_scale_score, decimals = 1)
```

This is the kind of table you'd put directly into a report someone is reading top to bottom — formatted, titled, and ready to present, rather than a raw printed data frame.

## Tables: DT for interactive, explorable tables

When the audience needs to search, sort, filter, or export the data themselves — rather than just read a fixed top-10 list — the `DT` package builds an interactive table widget, well suited to an HTML Quarto document:

```{r}
#| eval: false
library(DT)

static_file |>
  select(SchoolName, DistrictName, Subject, Performance, ScaleScore) |>
  head(250) |> 
  datatable(
    filter = "top",                 # column-level filter boxes
    extensions = "Buttons",
    options = list(
      scrollY = "400px",            # scroll instead of paginating
      paging = FALSE,
      dom = "Bfrtip",
      buttons = c("csv")            # a CSV download button, built in
    )
  )
```

A reader can filter to one school, sort by scale score, or export exactly the slice they care about — all without you building separate views for them. This is the table format you'll use in [Project Walkthrough 2](#sec-project-2).

## Inline values, tying it back to Quarto

As covered in [Quarto Essentials](#sec-quarto-essentials), any of the values you calculate in this chapter can be dropped directly into a sentence of prose using inline code:

```{r}
# The average ELA scale score across all schools in this sample is `r round(mean(static_file$ScaleScore[static_file$Subject == "ELA"], na.rm = TRUE), 1)`.
```

Combined with a `gt` table and a `ggplotly()` chart, this is the core toolkit behind both project walkthroughs that close out this book — starting with the next chapter.
