# Project Walkthrough 1: Automating a Multi-Tab Import {#sec-project-1}

This is the first of two end-to-end projects that pull together everything in this book so far. The scenario: you receive an Excel workbook with one tab per year of assessment data, and you need a clean, tidy, combined dataset — plus a validation file confirming the import looks right — produced automatically every time, rather than rebuilt by hand each cycle.

This belongs in a plain `.R` script rather than a `.qmd` document, since the goal is unattended automation, not a narrated report (see the file-type distinction in [Initial Setup](#sec-initial-setup)).

## The goal, end to end

1. Read every tab from a multi-year Excel workbook
2. Combine them into one tidy data frame
3. Reshape into long format for easier downstream analysis
4. Produce summary statistics as a validation check
5. Write both the tidy data and the validation summary to separate output folders

## Step 1: Read every tab

```{r}
library(tidyverse)
library(readxl)
library(janitor)

file_path <- "data/multi_year_assessment_results.xlsx"

sheet_names <- excel_sheets(file_path)

raw_list <- sheet_names |>
  map_dfr(~ read_excel(file_path, sheet = .x) |> mutate(source_sheet = .x))
```

Notice the addition of `mutate(source_sheet = .x)` inside the map step — tagging every row with which tab it came from before combining. This is cheap insurance: if anything looks wrong later, you can immediately trace a record back to its source tab rather than guessing.

## Step 2: Clean and combine

```{r}
tidy_data <- raw_list |>
  clean_names() |>
  filter(!is.na(scale_score))   # drop non-assessed records for this particular analysis
```

## Step 3: Reshape to long format

Suppose each year's tab had separate columns per subject (`ela_score`, `math_score`, `sci_score`) — a common real-world shape for a multi-year export. Reshape into one row per student per subject:

```{r}
tidy_long <- tidy_data |>
  pivot_longer(
    cols = ends_with("_score"),
    names_to = "subject",
    names_pattern = "(.*)_score",
    values_to = "scale_score"
  )
```

`names_pattern` here extracts just the subject portion of each column name (`ela`, `math`, `sci`) rather than keeping the full `_score` suffix — a small refinement on the basic `pivot_longer()` pattern from [Reshaping](#sec-reshaping).

## Step 4: Build a validation summary

Before trusting any pipeline's output, generate counts and summary statistics you can quickly eyeball against what you'd expect:

```{r}
validation_summary <- tidy_long |>
  group_by(source_sheet, subject) |>
  summarize(
    n_records      = n(),
    n_missing      = sum(is.na(scale_score)),
    avg_scale_score = mean(scale_score, na.rm = TRUE),
    min_score      = min(scale_score, na.rm = TRUE),
    max_score      = max(scale_score, na.rm = TRUE),
    .groups = "drop"
  )
```

This is the kind of table you'd glance at every time the pipeline runs — a sudden jump in `n_missing`, or an `avg_scale_score` outside the range you expect, is often the first sign something upstream changed before you'd otherwise notice.

## Step 5: Write outputs to separate subfolders

```{r}
dir.create("output/tidy", recursive = TRUE, showWarnings = FALSE)
dir.create("output/validation", recursive = TRUE, showWarnings = FALSE)

write_csv(tidy_long, "output/tidy/assessment_results_tidy.csv")
write_csv(validation_summary, "output/validation/validation_summary.csv")
```

`dir.create(..., showWarnings = FALSE)` quietly creates the folder if it doesn't exist yet, and does nothing (without throwing an error) if it already does — useful for a script meant to run unattended and repeatedly.

## Why this is worth the setup

The entire sequence above — reading, cleaning, reshaping, validating, and writing — is now a single script. Next year, when a new tab gets added to the source workbook, **nothing about this script needs to change.** `excel_sheets()` picks up the new tab automatically, `map_dfr()` combines it in automatically, and the validation summary will include it without you touching a single line. This is the compounding-returns argument for scripted work in its purest form: the investment happened once, here, and every future run is free.

The next chapter builds the second kind of deliverable — not an automated pipeline, but an interactive Quarto presentation meant for a human audience to actually read and explore.
