# Project Walkthrough 2: A Quarto Interactive Presentation {#sec-project-2}

The first project produced a clean, validated dataset behind the scenes. This one builds something a human being will actually open and read: an interactive Quarto presentation summarizing results, suitable for sharing with a principal, a board, or your own team.

## Setting up a new Quarto document

Start with a new `.qmd` file (see [Quarto Essentials](#sec-quarto-essentials) for the mechanics), with a YAML header set up for revealjs slides and self-contained sharing:


```{r}
# ---
# title: "School Performance Summary"
# subtitle: "Static File Sample — Current Year"
# format:
#   revealjs:
#     embed-resources: true
#     controls: true
# execute:
#   echo: false
#   warning: false
#   message: false
# ---
```


`embed-resources: true` matters here specifically because this file is meant to be shared — a single self-contained `.html` that opens correctly on anyone's machine, with no separate image folder that could go missing. `echo: false` at the document level means code won't show by default in any chunk unless you explicitly override it per-chunk — appropriate for a presentation meant for an audience, as opposed to the code-forward reference chapters earlier in this book.

## Load libraries and data once, near the top

```{r}
#| eval: false

library(tidyverse)
library(plotly)
library(gt)
library(DT)

static_file <- read_csv("data/static_file_sample.csv")
```

`include: false` hides both the code *and* any output from this setup chunk — appropriate here since loading libraries and data isn't something your audience needs to see, only something the rest of the document depends on.

## A slide with an interactive visualization


## Scale Score Distribution by Subject

```{r}
#| eval: false

p <- static_file |>
  filter(!is.na(ScaleScore)) |>
  ggplot(aes(x = ScaleScore, fill = Subject)) +
  geom_histogram(alpha = 0.7, position = "identity") +
  labs(x = "Scale Score", y = "Number of Students") +
  theme_minimal()

ggplotly(p)
```


Because this is wrapped in `ggplotly()` (from [Visualization](#sec-visualization)), anyone viewing the rendered slide can hover over bars for exact values, zoom into a score range, or toggle subjects on and off — interactivity a static image in a PDF simply can't offer.

## A slide with an interactive table


## Explore the Data

```{r}
#| eval: false
static_file |>
  filter(!is.na(Performance)) |>
  select(SchoolName, DistrictName, Subject, Performance, ScaleScore) |>
  datatable(filter = "top", options = list(pageLength = 8))
```

This single chunk hands your audience a live, searchable, filterable version of the underlying data, directly inside the presentation — rather than a static screenshot of a table they can't interact with.

## Using inline variables in your narrative

This is where Quarto's literate-programming approach (text and code woven together, from [Quarto Essentials](#sec-quarto-essentials)) does real work in a presentation context. Rather than typing a number that will quietly go stale the next time this file is regenerated against updated data:

`{{r}} 1+1`

```{r}
#| echo: false
#| include: true

# ``This year's sample includes `r n_distinct(static_file$SchoolName)` schools across `r n_distinct(static_file$DistrictName)` districts, with an overall ELA pass rate of `r scales::percent(mean(static_file$Passing[static_file$Subject == "ELA"], na.rm = TRUE))`.``

print("`r 1+1`")
```

Every number in that sentence is computed live from `static_file` at render time. Re-run this same document against next year's data, and every figure updates correctly with zero manual editing — directly delivering on the reproducibility argument that's been a throughline of this entire book.

## Previewing and sharing

Use **Preview** (`Ctrl/Cmd + Shift + K`, from [Quarto Essentials](#sec-quarto-essentials)) while you build, to check slides as you go. When you're ready to share:

1. Click **Preview** (or run `quarto render` from the terminal) to produce the final self-contained `.html` file
2. Click the pop-out button in the Viewer pane to open it in a full browser window, or simply double-click the rendered `.html` file from your file system — since `embed-resources: true` is set, it will open correctly anywhere, with no separate files required
3. Share that single `.html` file directly — email, shared drive, or anywhere else — and it will render identically on the recipient's end

## What you've actually built

A presentation that's simultaneously a finished deliverable *and* a script — one that can be re-rendered against next year's Static File with zero manual rework, that updates its own numbers automatically, and that gives your audience a genuinely interactive way to explore the underlying data rather than a static snapshot. This is, in miniature, the entire case for moving this kind of work into R: the investment is real, but it's a one-time investment that pays out every single time you run it again.
